Metric Spaces with Expensive Distances

### 2018-12-01 paper/draft.tex
Finished Section 5+Conclusion

\title{Metric Spaces with Expensive Distances}
\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{}{}
\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
\authorrunning{M. Kerber and A. Nigmetov}

@@

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams or Reeb graphs, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\R^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed,
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
\michael{This paragraph needs references. As you know the literature better than me,
I leave that part to you.}

@@

\section{Approximate Nearest Neighbors}
\label{sec:ann}
We consider the standard problem of finding an approximate nearest neighbor: given
$n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. This notation will be fixed throughout this 
section, and we shall also use the shorthand notation
\[
    r_i := \dist(p_i, q).
\]
We assume for simplicity
that all exact pairwise distances $\dist(p_i, p_j)$ are already computed
(a slight modification of the algorithm can also be applied if only a spanner is available).
Our goal is to reduce the number of computed distances $\dist(p_i, q)$. We shall also assume w.l.o.g.
that all $\dist(p_i, q)$ are distinct.

Our approach can be summarzied as follows. 
Fix a random permutation of the points of $P$ 
and consider the points in that order 
Fix a random permutation of the points of $P$ (to simplify notation,
we re-index them, so the order is again $p_1, \dots, p_n$).
During the loop, we maintain lower bounds of each $p_i$
to the query point $q$, which are initially all set to $0$.
We also remember the closest neighbor $c$ that we have seen so
far and its distance $v$ to $q$.
We refer to this point as the \emph{candidate}.
We maintain the invariant that $c$ is an approximate nearest neighbor
to $q$ for the points $\{p_1,\ldots,p_i\}$.
When reaching the point $p_i$, we check whether the lower
bound $a_i$ satisfies $a_i\geq \frac{v}{1+\eps}$.
If so, $c$ remains an approximate nearest neighbor and
we do not query the distance of $p_i$ to $q$.
Otherwise, we compute $\dist(p_i,q)$ and update the lower bounds
of all points according to the newly computed distance.
If $p_i$ is closer to $q$ than $c$, we update $c$ and $v$
accordingly. At the end of the loop, $c$ is an approximate nearest
neighbor. The pseudocode of the procedure follows.

@@

\begin{lemma}
\label{lem:bound_lemma}
Assume $r_i=\dist(p_i,q)$ is computed in the algorithm, and let $j>i$.
\begin{enumerate}
\item If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps}) r_i$, the algorithm will not compute the distance
of $p_j$ to $q$.
\item If $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$, the algorithm will not compute
the distance of $p_j$ to $q$.
\end{enumerate}
\end{lemma}
\begin{proof}
The algorithm computes $r_i$ by assumption and updates
all lower bounds. For $p_j$, it sets $a_j\gets \max (a_j,|\dist(p_i,p_j)-r_i|)$.
If $|\dist(p_i,p_j)-r_i|\geq (1+\frac{1}{1+\eps})r_i$,
so 
\[a_j\geq (1+\frac{1}{1+\eps}r_i - r_i = \frac{r_i}{1+\eps}.\]
Likewise, if $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$,
\[a_j\geq r_i-\dist(p_i,p_j) \geq r_i - \frac{\eps}{1+\eps}r_i = \frac{r_i}{1+\eps}.\]
In both cases, after the point $p_i$ is handled,
$v\leq r_i$ clearly holds. Since $v$ is only decreasing
and $a_j$ is only increasing in the algorithm,
it follows that $a_j\geq \frac{v}{1+\eps}$ when $p_j$ is handled,
so the algorithm proceeds without a distance computation.
\end{proof}

With the previous lemma, we can derive an upper bound for the number
of distance computations in an arbitrary subsequence of $p_1,\ldots,p_n$
as follows. In what follows, we let $c_1,\ldots,c_n$ denote the candidates
at the end of the $i$-th iteration of the loop, and $v_1,\ldots,v_n$
their distances to $q$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.

\begin{lemma}
Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
\[\left(\frac{2(2+\eps) v_k}{\eps v_\ell}\right)^{2d}\]
distances to $q$.
\end{lemma}
\begin{proof}
By the first part of Lemma~\ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_i=\frac{2+\eps}{1+\eps} v_k$
around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
is computed, the second part of Lemma~\ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma~\ref{lem:packing_lemma}.
\end{proof}

A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
the number of computed distances is a constant (since $v_k=v_\ell$). This means that to prove our statement, it would suffice to show
that the candidate changes only a logarithmic number of times in expectation.
While we have not found a simple proof for this claim, we can prove the statement with a slight variant of
that argument.

\begin{proof} (of Theorem~\ref{thm:ann_bound})
\michael{I need to rewrite this...to be continued asap}
Let $i_k$ be the {$k$-th} position where the sequence  $m_i := \min \Set{m_1, \dots, m_i}$
changes, so that
\[
m_1 = m_2 = \dots = m_{i_1 - 1} > m_{i_1} = m_{i_1 + 1} = \dots = m_{i_2 - 1} > m_{i_2} = \dots. 
\]
In other words, if we traversed the list $p_1, \dots, p_n$ and used the bruteforce algorithm
for exact nearest neighbor, then $i_k$ would the indices where the current candidate
changes, because we found a point that is closer to $q$ than all the previously seen points.
We shall refer to $i_k$ as the \textit{change position}.
Standard backwards analysis argument (e.g., \cite{dubch}, Sect. 4.4) shows that $k=O(log n)$.

    Consider two consecutive change positions $s:=i_j$ and $t:=i_{j+1}$. We claim that in the sequence
$p_s,...,p_{t-1}$
only a constant number of distances are computed by our algorithm, irrespective of its length. This claim immediately implies the theorem.

To see that it true, let $c_1,...,c_n$ be the candidates after the $i$-th step of
the for loop of the algorithm. We have no guarantee that $c_s=p_s$, because $p_s$ it is possible
    that $p_s$ was discarded without computing $r_s$ (see the picture). However, we know that
$d(c_s,q)<=(1+ \eps) r_s$.
Indeed, suppose that $c_s \neq p_s$ and $r_s < d(c_s,q)/(1+\eps)$. In this
    case $p_s$ cannot be discarded 
by its lower bound, and hence the distance to $p_s$ would
have been queried, but this distance is a minimum among $r_1, \dots, r_s$,
    and then we must reset $c_s$ to be $p_s$, a contradiction.

It follows that any point in the sequence $p_s,...,p_{t-1}$ for which a
distance is queried must lie in the ball around $c_s$ of radius at most $2(1+\eps)r_s$.

Moreover, we know that for any index $j=s,\dots, t-1$ it holds
$d(c_j,q)>=r_s$.

Indeed, assume the opposite. $c_j$ is equal to $p_k$ for some $k<t$. If $k<s$,
that is, the candidate came from before, than $p_s$ would not be in the
change position. If $s<k<t$, then at least one of the
elements in $p_{s+1},\dots,p_k$ is a change position, contradicting
the assumption that $p_s$ and $p_t$ were chosen to be consecutive change positions.

Hence, it follows that whenever a distance is queried, we can exclude a
ball of radius $\eps r_s$ around the query point. All queried points must
hence have pairwise distance of at least $ \eps r_s / 2$. 
Now we can apply the packing lemma \ref{lem:packing_lemma} to bound the number of
queried points by some constant that depends only on $\eps$ and the dimension. Indeed,
they all are contained in a ball of radius $R = 2(1+\eps)r_s$, and, if we divide this by $\eps r_s$, the ratio depends only on $\eps$.
\end{proof}

We should note that the constant factor that is obtained in this proof can be very pessimistic,
because in practice the algorithm can discard much more points.

%\begin{proof}
%    We start with the lemma of purely combinatorial (or probabilistic) nature.
%\begin{lemma}
%    Let $z$ be a random permutation of $n$ distinct real numbers,
%    $z = X_{1}, \dots, X_{n}$, and denote $x_t = \min \Set{X_1, \dots, X_t}$.
%    The expected number of positions $t$ where the sequence $X_t$ changes is $O(n)$.
%\end{lemma}
%    \begin{proof}
%        The proof is easily obtained with the standard backwards analysis technique and can be found in the appendix.
%        %Define random variable $Y_i$ to be 1, if $X_{i-1} = X_{i}$, and 0 otherwise.
%        %We must estimate $Y = \sum_{i = 2}^n Y_i$, and 
%        %\[
%        %    E[Y] = \sum_{i = 2}^n E[Y_i] = \sum_{i = 2}^n P(Y_i = 1).
%        %\]
%        %It remains to observe than $P(Y_i) = 1 / (n - i)$, so the expectation
%        %of $Y$ is the sum of the 
%    \end{proof}

%\begin{lemma}
%    There exists a constant $C$ that depends only on $\eps$ and the doubling dimension
%    such that after computing at most $C$ distances $\dist(p_k, q)$
%    either the algorithm verifies that the current candidate $p_i$ is a correct answer and terminates
%    or the algorithm finds another candidate $p_j$ such that $r_j < r_i$.
%\end{lemma}
%\begin{proof}
%    First, all points that are not contained in the ball $B(p_i, 2r_i)$
%    are discarded without computing the distance to them.
%    Now we have to examine all points $p_j$ inside this ball. 
%    Suppose that we cannot discard the current point $p_j$ based
%    on the lower bound for $r_j$, then we must compute the distance $r_j$.
%    There are two possibilities: either $r_j \leq r_i$ or $r_j > r_i$.
%    The latter case means that we discard the point $p_j$,
%    but we also discard all points of $P$ that are contained in some ball around it.
%    The triangle inequality tells us that we can take the ball of radius $r_j - r_i / (1 + \eps)$,
%    but in this proof it is sufficient to consider the ball of radius $r_i \eps / (1 + \eps)$.
%    In the figure this radius corresponds to the width of the annulus surrounding $q$,
%    and it is  geometrically clear that this ball is indeed smaller than the one we could have discarded.
%    The advantage of choosing this small radius is that, if we divide it by $2r_i$, we 
%    get the expression $\eps / (2(1 + \eps))$, that depends only on $\eps$.
%    By lemma \ref{lem:packing_lemma_dd} we know that there is an upper bound $T$ for the number of points
%    $p_i$ inside the ball $B(p_i, 2r)$ that are at the distance at least $r_i / 2 ( 1 + \eps)$
%    from each other, and the constant $T$ depends only on $\eps$ and the doubling dimension.
%    This means that even in the worst case after we have computed $T$ distances, we have discarded all
%    points of $P$ that lie outside of the ball $B(q, r_i)$, because all other points that lie in $B(p_i, 2r_i) \setminus B(q, r_i)$ 
%    do not require distance computation. Indeed, each of these points lies in
%    one of the balls $B(p_j, r_i \eps / (1 + \eps)$ for whose center we computed the distance already,
%    and is therefore discarded, Now, either we have exhausted the set $P$, and in this
%    case we terminate with an approximate nearest neighbor, or the next available $p_j$ that requires
%    distance computation necessarily lies in $B(q, r_i)$, thus we found a closer neighbor. This means that $C = T+1$.
%    Note that the algorithm still computes an approximate nearest neighbor, not the exact one, 
%    because some of the points that are inside the annulus can be discarded, 
%    if there is a point from $P$ close to them but outside of the annulus. 
%\end{proof}

%Now the theorem follows from the lemma by standard argument
%of the backwards analysis.

%\end{proof}

\subsection{Experimental Results}

@@

\section{Conclusion and Future Work}
\label{sec:conclusion}
We presented experimental evidence showing that in low
doubling dimension we can avoid many distance computations,
if we try to get as much information as possible from the triangle inequality
applied to the previously computed distance. The \textsc{BlindGreedy} spanner
performs especially well, being able to produce  sparse spanners even for small values of  $\eps$.
For the approximate nearest neighbor algorithm we proved a logarithmic upper bound,
and, even though the constant in the proof was very large, we demonstrated
experimentally that this does not show in practice.

The next step on the theoretical side could be proving some upper bound
for blind spanners; it is also interesting to look at other metric space
algorithms and data structures in the setting of expensive distance computation.
On the more applied side, the further development of this approach would
be to find finite spaces of low dimension in practically
important cases, like spaces of persistence diagrams or images endowed with Wasserstein (a.k.a. Earth Movers)
metric. We know that low-dimensional subspaces of functions do not necessarily produce
low-dimensional spaces of persistence diagrams, but there may be some conditions
which guarantee that dimension does not increase too much.
% An easy example: consider a fixed function $f$ to which we apply
% a finite-dimensional family of perturbations, $f(x) + g(x; \lambda_1, \dots, \lambda_n)$;
% for instance, each $\lambda_i$ is associated with some $x_i$ and adds a Gaussian noise
% scaled by $\lambda$.
% If the perturbations are very small, then they will produce some small wiggles on the function,
% and then the optimal matcih

>>>>

\title{Metric Spaces with Expensive Distances}
\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
\authorrunning{M. Kerber and A. Nigmetov}

@@

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams or Reeb graphs, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\R^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed,
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric.
\michael{This paragraph needs references. As you know the literature better than me,
I leave that part to you.}

@@

\section{Approximate nearest neighbors}
\label{sec:ann}
We consider the standard problem of finding an approximate nearest neighbor: given
$n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. This notation will be fixed throughout this 
section, and we shall also use the shorthand notation
\[
    r_i := \dist(p_i, q).
\]
We assume for simplicity
that all exact pairwise distances $\dist(p_i, p_j)$ are already computed
(a slight modification of the algorithm can also be applied if only a spanner is available).
Our goal is to reduce the number of computed distances $\dist(p_i, q)$.

Our approach can be summarzied as follows. 
Fix a random permutation of the points of $P$ 
and consider the points in that order 
(to simplify notation,
we re-index them, so the order is again $p_1, \dots, p_n$).
During the loop, we maintain lower bounds of each $p_i$
to the query point $q$, which are initially all set to $0$.
We also remember the closest neighbor $c$ that we have seen so
far and its distance $v$ to $q$.
We refer to the point $c$ as the \emph{candidate}.
We maintain the invariant that $c$ is an approximate nearest neighbor
to $q$ for the points $\{p_1,\ldots,p_i\}$.
When reaching the point $p_i$, we check whether the lower
bound $a_i$ satisfies $a_i\geq \frac{v}{1+\eps}$.
If so, $c$ remains an approximate nearest neighbor and
we do not query the distance of $p_i$ to $q$.
Otherwise, we compute $\dist(p_i,q)$ and update the lower bounds
of all points according to the newly computed distance.
If $p_i$ is closer to $q$ than $c$, we update $c$ and $v$
accordingly. At the end of the loop, $c$ is an approximate nearest
neighbor. The pseudocode of the procedure follows.

@@

\begin{lemma}
\label{lem:bound_lemma}
Assume $r_i=\dist(p_i,q)$ is computed in the algorithm, and let $j>i$.
\begin{enumerate}
\item If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps}) r_i$, the algorithm will not compute the distance
of $p_j$ to $q$.
\item If $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$, the algorithm will not compute
the distance of $p_j$ to $q$.
\end{enumerate}
\end{lemma}
\begin{proof}
The algorithm computes $r_i$ by assumption and updates
all lower bounds. For $p_j$, it sets $a_j\gets \max (a_j,|\dist(p_i,p_j)-r_i|)$.
If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps})r_i$, it follows that 
\[a_j\geq (1+\frac{1}{1+\eps})r_i - r_i = \frac{r_i}{1+\eps}.\]
Likewise, if $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$,
\[a_j\geq r_i-\dist(p_i,p_j) \geq r_i - \frac{\eps}{1+\eps}r_i = \frac{r_i}{1+\eps}.\]
In both cases, after the point $p_i$ is handled,
$v\leq r_i$ clearly holds. Since $v$ is only decreasing
and $a_j$ is only increasing in the algorithm,
it follows that $a_j\geq \frac{v}{1+\eps}$ when $p_j$ is handled,
so the algorithm proceeds without a distance computation.
\end{proof}

In what follows, we let $c_1,\ldots,c_n$ denote the candidates
at the end of the $i$-th iteration of the loop, and $v_1,\ldots,v_n$
their distances to $q$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
With the previous lemma, we can derive an upper bound for the number
of distance computations in an arbitrary subsequence of $p_1,\ldots,p_n$
as follows.

\begin{lemma}
\label{lem:sequence_lemma}
Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
\[\left(\frac{2(2+\eps) v_k}{\eps v_\ell}\right)^{2d}\]
distances to $q$.
\end{lemma}
\begin{proof}
By the first part of Lemma~\ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
is computed, the second part of Lemma~\ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma~\ref{lem:packing_lemma}.
\end{proof}

A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
the number of computed distances is a constant (since $v_k=v_\ell$). 
This means that to prove Theorem~\ref{thm:ann_bound}, it would suffice to show
that the candidate changes only a logarithmic number of times in expectation.
While we have not found a simple proof for this claim, we can prove the statement with a slight variant of
that argument.

\begin{proof} (of Theorem~\ref{thm:ann_bound})
%
In the sequence $p_1,\ldots,p_n$, let $p_k$ be a point 
such that $r_i<r_k$ for all $1\leq i\leq k-1$. We call an element
of this form a \emph{minimum} of the sequence. 
A standard backwards analysis argument~\cite{seidel-backwards} shows that the probability
of $p_k$ being a minimum is at most $1/k$, so that the number of minima
in the sequence is $O(\log n)$ in expectation.

Note that for $\eps>0$,
a minimum $p_k$ is not necessarily the candidate $c_k$ because a previous point
in the sequence close to $p_k$ might have caused the lower bound $a_k$ to be
in the interval $[\frac{v_k}{1+\eps},v_k]$, which leads to not
computing the distance $r_k$. However, it is true that
$v_k\leq (1+\eps)r_k$, because otherwise, $c_k$ would not be an approximate
nearest neighbor of $\{p_1,\ldots,p_k\}$.

Now, let $p_k$, $p_\ell$ be two consecutive minima in the sequence
(we also allow that $\ell=n+1$ if $k$ is the last minimum in the sequence).
Note that $v_{\ell-1}\geq r_k$ because each $v_j$ is equal to $r_i$ for
some $i\leq j$, and in the sequence $r_1,\ldots,r_{\ell-1}$, $r_k$
is minimal by construction. Using Lemma~\ref{lem:sequence_lemma},
the number of distance computations among the points
$p_k,\ldots,p_{\ell-1}$ is at most
\[
\left(\frac{2(2+\eps) v_k}{\eps v_{\ell-1}}\right)^{2d}\leq \left(\frac{2(2+\eps)(1+\eps)r_k}{\eps r_k}\right)^{2d}=\left(\frac{2(2+\eps)(1+\eps)}{\eps}\right)^{2d},\]
which is a constant depending only of $\eps$ and $d$, irrespective of the length of the sequence.
Since $p_1,\ldots,p_n$ decomposes into $O(\log n)$ such sequences in expectation, the result follows.
\end{proof}

We point out that the proof fails for $\eps=0$ because in that case, we cannot exlude an $\eps$-ball
of close-by points as in the second part of Lemma~\ref{lem:bound_lemma}, and the packing argument fails.
Indeed, as the example in Figure~\ref{} shows, there are point sets where the expected number
of distance computations for exact nearest neighbor is linear.

Finally, we remark that a fast $2$-approximation algorithm for $\dist$ would lead to a straight-forward
optimization: compute a $2$-approximation of $\dist(p_i,q)$ for all $1\leq i\leq n$
and let $m$ denote the minimal approximate distance encountered. Then, we can discard all points
whose approximate distance is larger than $2m$, and run the above algorithm on the remaining points.

\subsection{Experiments on approximate nearest neighbors}

@@

\section{Conclusion and future work}
\label{sec:conclusion}
%
We have introduced a new cost model for the analysis of algorithms
for metric spaces that fits the situation that computing an individual distance
is more costly than other types of primitive operations.
Our theoretical and experimental results are under the usual assumption
that the metric space has a low doubling dimension.
However, in our motivating example of collections of persistence diagrams
or Reeb graphs, this assumption does not hold. For instance,
the space of persistence diagrams has an infinite doubling dimension.
Nevertheless, realisitic data sets are usually not just a random sample
in that infinite-dimensional space, but have structures
(e.g. clusters of close-by diagrams) which should be favorable for our approach
We plan to consider the quality of our algorithms for persistence diagrams
as future work.

On the theoretical side, the obvious next question is whether our strategy
for blind spanners yields a linear spanner in expectation. 
Our experiments are not conclusive enough in this respect to make this conjecture yet.
However, it has been brought to our attention\footnote{Yusu Wang, personal communication}
that the size of the blind spanner is bounded by the \emph{weight} of the WSPD
which is the sum of the cardinalities of all pairs in a WSPD.
The weight of a WSPD can be quadratic, but premilinary
experimental evaluation on worst-case examples do not show such a quadratic
behavior. Therefore, we postpone the theortical analysis of the spanner construction
to an extended version of this article.

The existence of a $2$-approximation algorithm did not help us to significantly
reduce the number of exact distance computations, although it seems obvious
that knowing the all approximate distances is useful.
We pose the question what heuristic could make more use of this feature.

\ignore{
We presented experimental evidence showing that in low
doubling dimension we can avoid many distance computations,
if we try to get as much information as possible from the triangle inequality
applied to the previously computed distance. The \textsc{BlindGreedy} spanner
performs especially well, being able to produce  sparse spanners even for small values of  $\eps$.
For the approximate nearest neighbor algorithm we proved a logarithmic upper bound,
and, even though the constant in the proof was very large, we demonstrated
experimentally that this does not show in practice.

The next step on the theoretical side could be proving some upper bound
for blind spanners; it is also interesting to look at other metric space
algorithms and data structures in the setting of expensive distance computation.
On the more applied side, the further development of this approach would
be to find finite spaces of low dimension in practically
important cases, like spaces of persistence diagrams or images endowed with Wasserstein (a.k.a. Earth Movers)
metric. We know that low-dimensional subspaces of functions do not necessarily produce
low-dimensional spaces of persistence diagrams, but there may be some conditions
which guarantee that dimension does not increase too much.
% An easy example: consider a fixed function $f$ to which we apply
% a finite-dimensional family of perturbations, $f(x) + g(x; \lambda_1, \dots, \lambda_n)$;
% for instance, each $\lambda_i$ is associated with some $x_i$ and adds a Gaussian noise
% scaled by $\lambda$.
% If the perturbations are very small, then they will produce some small wiggles on the function,
% and then the optimal matcih
}
